Informing policy via dynamic models: Cholera in Haiti

### 2024-02-29 edits.tex
discussion edits

% \newcommand\editDiscussionOne{Model diagnostics are instrumental in enhancing model fits and identifying limitations of models that cannot readily be fixed. In our case study, we compared the Models~1--3 of \cite{lee20} to simple statistical benchmarks, enabling us to identify deficiencies in the models that could be improved. In the case of Model~3, comparing a fitted model to a benchmark and then following up by questioning why the model did not describe the data as efficiently as a simple statistical model helped us discern that the model fell short in explaining the surge in cholera transmission following Hurricane Matthew. This led us to suggest a model modification that enchanced model fit. When a mechanistic model beats a statistical benchmark, we can begin to critically evalutate the causal implications of the fitted model, carefully acknowledging the fact that beating a simple benchmark alone does not imply that the model is both a useful quantitative and qualitative description of the data. In Model~1, we found that the flexible seasonality term closely mirrors the seasonal rainfall patterns in Haiti, reaffirming alternative evidence of the importance of rainfall as a driver of cholera infection and increasing our confidence in Model~1's ability to capture the critical dynamics of the cholera outbreak.}
% 
\newcommand\editDiscussionOne{Model diagnostics are instrumental for improving model fit and exposing unresolvable limitations. In our case study, we compared the three models from Lee et al. (2020) to statistical benchmarks, revealing areas for enhancement. For example, comparisons of Model~3 to a benchmark revealed its inadequacy in accounting for the post-hurricane increase in transmission, leading to a beneficial model refinement. When a mechanistic model beats a statistical benchmark, we can begin to critically evalutate causal implications, carefully acknowledging that beating a simple benchmark alone does not imply the model is a useful qualitative description of the data. The recalibrated version of Model~1 outperforms its benchmark, enabling evaluation of causal implications. When doing so, we find that the fitted model provides a causal description of the dynamic system that is consitent with known features of the system, such as the importance of rainfall as a driver of cholera infection. The congruency between causal implications of the model and our belief about the dynamic system, coupled with a strong quantitative description of obsered data relative to a benchmark, increases our confidence that the model adequately captures critical dynamics of the system.}

@@

% \newcommand\editDiscussionThree{Another important consideration when fitting a mechanistic model to a dynamic system is the complexity of the model. Mathematical models necessarily simplify complex dynamical systems in order to make inference possible. Decisions about which features of a system to incorporate in a model and which should be excluded can significantly influence the model-based conclusions. However, these decisions can be tested using a likelihood-based framework. Standard model decisions cover deterministic versus stochastic modeling and the decision to include or exclude a meta-population structure. Model benchmarks can assist in determining the level of complexity to be used in the model, enabling researchers to gauge a model's relative performance against a better-understood statistical model.}
% 
\newcommand\editDiscussionTwo{When fitting a mechanistic model to a dynamic system, the complexity of the model warrants careful attention. Mathematical models simplify complex dynamical systems, highlighting essential features that both facilitate scientific understanding and enable inference. In this simplification process, careful consideration must be given to which facets of the system are incorporated into the model and which are omitted, as these choices can shape the conclusions drawn from the model. These modeling decisions are driven both by the need for tractable inference and by the pursuit of scientific parsimony. Standard modeling choices—such as adopting a deterministic or stochastic approach, or incorporating a meta-population structure—reflect the balance between realism and simplicity.} %Benchmarking models against established and simpler statistical models in a likelihood based framework can help researchers determine an appropriate level of complexity, ensuring that models are both manageable and meaningful in their portrayal of the underlying dynamics.}

\newcommand\editDiscussionThree{Along with enabling comparisons to statistical benchmarks, likelihood-based methods aid in determining an appropriate level of model complexity. Models fit to the same data can be compared using a criteria such as AIC. Nested model variations are particularly useful as they enable formal statistical testing of the nested features. We illustrated the benefits of examining nested model features, enhancing all three models. Model~1 included a time-varying transmission rate; Model~2 assessed a phase-shift parameter in seasonal cholera peaks; Model~3 incorporated hurricane-related parameters. Each addition received strong statistical support over their simpler predecessors.}

\newcommand\editDiscussionFour{Unmodeled features of a dynamic system can lead to spurious or misleading parameter estimates if the feature greatly impacts observed data. This is because the model must compensate for these unaccounted features using its inherent flexibility. In deterministic models, parameter estimation is the sole source of such flexibility. When the influence of unmodeled dynamics is substantial, it can skew parameter estimates toward scientifically implausible values, a phenomenon observed in the estimated parameter values of Model~2. Incorporating demographic and environmental stochasticity into models can mitigate the impact of these unmodeled features. These stochastic phenomena are not only arguably present in biological systems, but their inclusion in a model also allows observed data variations to be attributed to inherent randomness rather than to distorted parameter values. Incorporating extra-demographic stochasticity in a model has also been shown to be beneficial \cite{he10,stocks20,li23}, which is consistent with our results: Models~1 and 3 suggest the presence of extra-demographic stochsticity, as evidenced by the confidence intervals provided in Suplement~S8.}

\newcommand\editDiscussionFive{The three models in this case study consider a range of possible model complexity. Of these models, Model~3 is distinct in that it is both stochastic and has a meta-population structure, making it challenging to draw likelihood-based inferences. In this paper, we demonstrated how this model class could be calibrated to incidence data using the innovative IBPF algorithm. One of only a few examples of fitting a model with this level of complexity via maximum likelihood, this case study exemplifies the algorithm's potential benefits and provides an example for future researchers on a possible approach to fitting a high-dimensional non-linear model.}

\newcommand\editDiscussionSix{If forecasts are an important component of a modeling task, the forecasts should be consistent with the available data, particularly at the most recently available time points. In our case study, we do this by simulating forward from the filtering distribution, as this proceedure conditions latent variables on the available data. This type of forecasting, however, is not possible using a deterministic model, where only initial conditions and parameter values can influence forecasts. Deterministic simulations may also result in over-confidence in model forecasts, as they can only account for uncertainty due to the parameter estimation proceedure \cite{king15}. Despite their limitations, deterministic models can offer valuable insights into dynamic systems. As with all models, researchers should consider and acknoweledge these limitations when applying deterministic models and interpreting their outcomes.}

>>>>

% \newcommand\editDiscussionOne{Model diagnostics are instrumental in enhancing model fits and identifying limitations of models that cannot readily be fixed. In our case study, we compared the Models~1--3 of \cite{lee20} to simple statistical benchmarks, enabling us to identify deficiencies in the models that could be improved. In the case of Model~3, comparing a fitted model to a benchmark and then following up by questioning why the model did not describe the data as efficiently as a simple statistical model helped us discern that the model fell short in explaining the surge in cholera transmission following Hurricane Matthew. This led us to suggest a model modification that enchanced model fit. When a mechanistic model beats a statistical benchmark, we can begin to critically evalutate the causal implications of the fitted model, carefully acknowledging the fact that beating a simple benchmark alone does not imply that the model is both a useful quantitative and qualitative description of the data. In Model~1, we found that the flexible seasonality term closely mirrors the seasonal rainfall patterns in Haiti, reaffirming alternative evidence of the importance of rainfall as a driver of cholera infection and increasing our confidence in Model~1's ability to capture the critical dynamics of the cholera outbreak.}
% 
\newcommand\editDiscussionOne{Model diagnostics are a key tool for exposing unresolved model limitations and improving model fit. In our case study, we compared the three models from Lee et al. (2020) to statistical benchmarks, revealing areas for enhancement. For example, comparisons of Model~3 to a benchmark revealed its inadequacy in accounting for the post-hurricane increase in transmission, leading to a beneficial model refinement. When a mechanistic model is competitive with statistical benchmarks, we have a license to begin critical evaluation of its causal implications.
If a model falls far behind simple benchmarks, there is likely to be substantial limitation in the data analysis that should be identified and remedied.
In our case study, the recalibrated version of Model~1 outperforms its benchmark, enabling evaluation of causal implications. When doing so, we find that the fitted model provides a causal description of the dynamic system that is consistent with known features of the system, such as the importance of rainfall as a driver of cholera infection. The congruency between causal implications of the model and our belief about the dynamic system, coupled with a strong quantitative description of observed data relative to a benchmark, provides support for viewing the model as a plausible quantitative representation of the system under investigation.}

@@

% \newcommand\editDiscussionThree{Another important consideration when fitting a mechanistic model to a dynamic system is the complexity of the model. Mathematical models necessarily simplify complex dynamical systems in order to make inference possible. Decisions about which features of a system to incorporate in a model and which should be excluded can significantly influence the model-based conclusions. However, these decisions can be tested using a likelihood-based framework. Standard model decisions cover deterministic versus stochastic modeling and the decision to include or exclude a meta-population structure. Model benchmarks can assist in determining the level of complexity to be used in the model, enabling researchers to gauge a model's relative performance against a better-understood statistical model.}
% 
\newcommand\editDiscussionTwo{When fitting a mechanistic model to a dynamic system, the complexity of the model warrants consideration. Mathematical models provide simplified representations of complex systems, with the simplicity serving both to facilitate scientific understanding and to enable statistical inference on unknown parameters.
In our case study, employing a deterministic dynamic model was found to be an over-simplification by comparing model fit with benchmarks and stochastic models.
Including stochastic metapopulation dynamics permitted investigation of spatiotemporal data, the main cost being additional complexity of numerical computations.}
%In this simplification process, careful consideration must be given to which facets of the system are incorporated into the model and which are omitted, as these choices can shape the conclusions drawn from the model. These modeling decisions are driven both by the need for tractable inference and by the pursuit of scientific parsimony.
%Standard modeling choices—such as adopting a deterministic or stochastic approach, or incorporating a meta-population structure—reflect the balance between realism and simplicity.} %Benchmarking models against established and simpler statistical models in a likelihood based framework can help researchers determine an appropriate level of complexity, ensuring that models are both manageable and meaningful in their portrayal of the underlying dynamics.}

\newcommand\editDiscussionThree{Along with enabling comparisons to statistical benchmarks, likelihood-based methods aid in determining an appropriate level of model complexity. Models fit to the same data can be compared using a criteria such as AIC. Nested model variations are particularly useful as they enable formal statistical testing of the nested features via likelihood ratio tests. Our case study demonstrated the examination of nested model features for all three models. Model~1 investigated a time-varying transmission rate; Model~2 assessed a phase-shift parameter in seasonal cholera peaks; Model~3 incorporated hurricane-related parameters.}

\newcommand\editDiscussionFour{Unmodeled features of a dynamic system can lead to spurious or misleading parameter estimates if the features substantially impact observed data. This is because the model must compensate for these unaccounted features using its inherent flexibility. In deterministic models, the choice of parameters is the sole source of such flexibility.
For our case study, some of the parameter estimates for the deterministic Model~2 are implausible [WHICH ONES?] and this may be explained by compensation for model misspecification.
Incorporating demographic and environmental stochasticity into models can mitigate the impact of unmodeled features. Stochastic phenomena are not only arguably present in biological systems, but their inclusion in a model also allows observed data variations to be attributed to inherent uncertainty rather than to distorted parameter values.
Models~1 and 3 suggest the presence of extra-demographic stochsticity \cite{he10,stocks20,li23}, as evidenced by the confidence intervals provided in Supplement~S8.
}

\newcommand\editDiscussionFive{The three models in this case study consider a range of possible model complexity. Of these models, Model~3 is distinct in that it is both stochastic and has a meta-population structure, making it challenging to draw likelihood-based inferences. In this paper, we demonstrated how this model class could be calibrated to incidence data using the innovative IBPF algorithm. One of only a few examples of fitting a model with this level of complexity via maximum likelihood, this case study exemplifies the algorithm's potential benefits and provides an example for future researchers on a possible approach to fitting a high-dimensional non-linear model.
THIS PARAGRAPH SEEMS TO OVERLAP OTHERS. MAYBE FIT WHATEVER IS NEEDED INTO OTHER PARAGRAPHS AND DELETE?}

\newcommand\editDiscussionSix{If forecasts are an important component of a modeling task, the forecasts should be consistent with the available data, particularly at the most recently available time points. In our case study, we do this by simulating forward from the filtering distribution, as this proceedure conditions latent variables on the available data. This type of forecasting, however, is not possible using a deterministic model, where only initial conditions and parameter values can influence forecasts. Deterministic simulations may also result in over-confidence in model forecasts, as they can only account for uncertainty due to the parameter estimation proceedure \cite{king15}. Despite their limitations, deterministic models can offer valuable insights into dynamic systems. As with all models, researchers should consider and acknoweledge these limitations when applying deterministic models and interpreting their outcomes.
I WONDER WHETHER PEOPLE SERIOUSLY FORECASTING WITH DETERMINISTIC MODELS HAVE OTHER TRICKS. FOR EXAMPLE, WITH WEATHER FORECASTING MODELS, THEY ARE OFTEN WRITTEN AS DETERMINISTIC BUT PEOPLE DO SOMETHING ENKF-LIKE ANYHOW. JEFF SHAMAN HAS ALSO DONE THINGS LIKE THIS WITH EPI MODELS. ANOTHER OPTION COULD BE TO FIT IN A MOVING WINDOW. ANOTHER THING: THE PENULTIMATE SENTENCE MAKES A FAIRLY STRONG CLAIM THAT IS NOT SUPPORTED BY EVIDENCE HERE. PERHAPS WE CAN REMIND READERS THAT DETERMINISTIC MODELS REMAIN WIDELY USED, BUT IT WOULD BE NICE TO CLARIFY WHAT THEY ARE GOOD FOR SINCE WE SPEND SO MUCH TIME ON THEIR LIMITATIONS.}

### 2024-02-29 ms.Rnw
discussion edits

The ongoing global COVID-19 pandemic has \old{provided a clear example on} \new{demonstrated} how government policy may be affected by the \old{conclusions of scientific models} \new{inferences drawn from mathematical modeling} \cite{saltelli20}.
\old{This article demonstrates that fitting appropriate scientific models to guide policy is a challenging statistical task.}
\new{However, developing credible models---which are supported by data and can lend insights into a dynamic system---remains a challenging task.
In this article, we demonstrate that increasing the current standards of mathematical modeling of ecological systems can increase our capacity to make sense of a complex, nonlinear system's dynamics and make data-driven inferences.
We suggest devoting more effort to assess dynamic models intended for system inference.
Explicitly, we urge researchers to search for indications in available data that quantify the reliability of model-based inferences.
We illustrate that this can be achieved by comparing fitted models to a suitable statistical benchmark and verifying that model outcomes---including estimated parameters, latent states, and seasonal trends---are justified in the system context.}

>>>>

The ongoing global COVID-19 pandemic has \old{provided a clear example on} \new{demonstrated} how government policy may be affected by the \old{conclusions of scientific models} \new{inferences drawn from mathematical modeling} \cite{saltelli20}.
\old{This article demonstrates that fitting appropriate scientific models to guide policy is a challenging statistical task.}
\new{However, the development of credible models---which are supported by data and can lend quantitative insights into a dynamic system---remains a challenging task.
In this article, we demonstrate opportunites available for raising the current standards of statistical inference for mathematical models of biological systems.

We present methodology consistent with existing guidelines but exceeding standard practice.  
In particular, we show the value of comparing the likelihood of fitted mechanistic models versus non-mechanistic benchmarks.
Further, we use likelihood-based diagnostic methods to identify specific limitations of proposed models.
We use likelihood profile methods to assess parameter identifiability and to construct confidence intervals for parameter estimates. 
We assess the accuracy of numerical approximations involved in our calculations.
When reaching conclusions, we identify potential consequences of confounded variables and model misspecification.
}
